---
classoption: "runningheads"
title: "CytokineFinder: identifying disease associated cytokines"
shorttitle: ""
# If the paper title is too long for the 
# running head, you can set an abbreviated 
# paper title here.
# shorttitle: "Abbreviated paper title"
authors: 
  - name: "Jeffrey Tang"
    inst: "1,2"
    orcid: "0000-1111-2222-3333"
  - name: "Amrit Singh"
    inst: "1,2"
    orcid: "1111-2222-3333-4444"
# First names are abbreviated in the running head.
# If there are more than two authors, 'et al.' is used.
authorrunning: "F. Author et al." 
institutes: 
  - name: "Department of Anesthesiology, Pharmacology & Therapeutics, The University of British Columbia, Vanouver, Canada"
  - name: "Centre for Heart Lung Innvation, St. Paul's Hosital, Vanouver, Canada"
abstract: | 
  The abstract should briefly summarize the contents of the paper in 150--250 words.
keywords: 
  - First keyword
  - Second keyword
  - Another keyword
bibliography: lncs-test.bib
# Use csl when using default citation_package for Pandoc citeproc
csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
# citations with arabic numbers is prefered when using natbib
natbiboptions: numbers

output: 
  rticles::lncs_article:
    citation_package: default # or natbib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.extra = "")  # This forces knitr to label all figures.
```

# Introduction

- importance of identifying driver cytokines
- mention limitations in publicly available cytokine data
- state hypothesis: using gene expression data we can identify the driver cytokines of disease

# Methods
## public cytokine data curation
  - cytokine stimulations
  - anti-cytokine treatment
  - studies with paired gene-expression and protein data
  
## study data
  - asthma gene expression data
  - validate with cytokien protein levels
  
## cytokneFinder
  - add workflow diagram of study (figure 1)
  - briefly describe each method; but add more details of each method in the supplement
  
## benchmarking analysis

## data analysis of study data
  - apply best method to study data

# Results

## Benchmarking (figure 2)
  - impact of annotations
  - which method+ann works best

## study results
   - patient demographics
   - cytokine finder results with study data
   - hypothesis testing for significant cytokines

# Discussion



## A Subsection Sample

Please note that the first paragraph of a section or subsection is not indented. 
The first paragraph that follows a table, figure, equation etc. does not need an indent, either.

Subsequent paragraphs, however, are indented.

### Sample Heading (Third Level)

Only two levels of headings should be numbered. 
Lower level headings remain unnumbered; they are formatted as run-in headings.

#### Sample Heading (Fourth Level)

The contribution should contain no more than four levels of headings. 
Table\ \ref{tab:table} gives a summary of all heading levels.

```{r table}
knitr::kable(head(pressure), 
             format = "latex",
             caption = "Table captions should be placed above the tables.")
```

\noindent Displayed equations are centered and set on a separate line.

\begin{equation}
x + y = z
\end{equation}

Please try to avoid rasterized images for line-art diagrams and schemas. 
Whenever possible, use vector graphics instead (see Fig.\ \ref{fig:fig1}).

```{r fig1, fig.cap = "A figure caption is always placed below the illustration. Please note that short captions are centered, while long ones are justified by the macro package automatically."}
plot(pressure)
```

::: {.theorem latex=true}
This is a sample theorem. The run-in heading is set in bold, while the following text appears in italics. Definitions, lemmas, propositions, and corollaries are styled the same way.
:::

The environments 'definition', 'lemma', 'proposition', 'corollary', 'remark', and 'example' are defined in the LLNCS documentclass as well.

::: {.proof latex=true}
Proofs, examples, and remarks have the initial word in italics, while the following text appears in normal font.
:::

For citations of references, use [@Nuncio2011].
Multiple citations are grouped [@Levitus2012], [@Raphael2004].

### Acknowledgements
Please place your acknowledgments at the end of the paper, preceded by an unnumbered run-in heading (i.e. 3rd-level heading).

# References {#references}