---
title: "LRI Database Curation for Benchmarking"
author: "Jeffrey Tang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LRI Database Curation for Benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r default-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(cytokineFindeR)
```

This vignette documents how the bundled `dbs_cytosig` object is derived from the
raw LRI archive `dbs_all`. The goal is to restrict each database's
ligand-receptor sets to the cytokines covered by CytoSig, so that every method in
the benchmark scores the same panel of ligands.

# Load the LRI databases

`dbs_all` is a list of databases; each database is itself a named list mapping a
ligand to its receptor genes.

```{r load-db}
data("dbs_all")
names(dbs_all)
```


## A CytoSig-specific cytokine-to-gene mapping

For benchmarking we keep only ligands present in the CytoSig panel. The mapping
below translates each CytoSig cytokine label to its gene symbol(s); some
cytokines are multi-subunit and therefore map to several genes.

```{r list-cytokines}
# Cytokine-to-gene mapping (CytoSig panel only)
cytokine_to_gene_mapping <- list(
  "INHBA" = "INHBA",
  "BDNF" = "BDNF",
  "BMP2" = "BMP2",
  "BMP4" = "BMP4",
  "BMP6" = "BMP6",
  "BMP7" = "BMP7",
  "CCL2" = "CCL2",
  "CXCL12" = "CXCL12",
  "EGF" = "EGF",
  "EPO" = "EPO",
  "FGF2" = "FGF2",
  "GDF11" = "GDF11",
  "HGF" = "HGF",
  "HMGB1" = "HMGB1",
  "IFNG" = "IFNG",
  "IGF1" = "IGF1",
  "IL10" = "IL10",
  "IL13" = "IL13",
  "IL15" = "IL15",
  "IL17A" = "IL17A",
  "IL18" = "IL18",
  "IL1A" = "IL1A",
  "IL1B" = "IL1B",
  "IL2" = "IL2",
  "IL21" = "IL21",
  "IL22" = "IL22",
  "IL3" = "IL3",
  "IL33" = "IL33",
  "IL4" = "IL4",
  "IL6" = "IL6",
  "INS" = "INS",
  "LIF" = "LIF",
  "LTA" = "LTA",
  "NRG1" = "NRG1",
  "OSM" = "OSM",
  "PDGFB" = "PDGFB",
  "PDGFD" = "PDGFD",
  "TGFA" = "TGFA",
  "TGFB1" = "TGFB1",
  "TGFB3" = "TGFB3",
  "TSLP" = "TSLP",
  "VEGFA" = "VEGFA",
  "WNT3A" = "WNT3A",
  "WNT5A" = "WNT5A",

  # Single-gene conversions (name changes)
  "CD40L" = "CD40LG",
  "GCSF" = "CSF3",
  "GMCSF" = "CSF2",
  "IFN1" = "IFNA1",
  "IFNL" = "IFNL1",
  "MCSF" = "CSF1",
  "TNFA" = "TNF",
  "TRAIL" = "TNFSF10",
  "TWEAK" = "TNFSF12",
  "PGE2" = "PTGS2",

  # Multi-subunit cytokines (only those in the CytoSig list)
  "IL12" = c("IL12A", "IL12B"),
  "IL27" = c("IL27A", "EBI3"),
  "IL36" = c("IL36A", "IL36B", "IL36G")
)
```

## Helper: reduce each LRI database to CytoSig features

The helper below walks every database and, for each CytoSig cytokine, gathers the
union of receptor genes across that cytokine's gene symbol(s). Self-references
(where a receptor entry equals the cytokine name itself) are removed to avoid a
ligand scoring its own expression.

```{r helper-fun}
# Build CytoSig-only features across databases
create_cytokine_dbs <- function(dbs_all, cytokine_mapping) {
  # For each database
  lapply(dbs_all, function(db) {
    # For each cytokine in the mapping
    cytokine_features <- lapply(names(cytokine_mapping), function(cytokine) {
      genes <- cytokine_mapping[[cytokine]]

      # Get all receptors for these genes (ignore missing genes)
      all_receptors <- unlist(db[genes[genes %in% names(db)]])
      unique_receptors <- unique(all_receptors[!is.na(all_receptors)])

      # Remove only exact cytokine-name matches (keep subunit interactions)
      unique_receptors[unique_receptors != cytokine]
    })

    # Name the list and drop empty features
    names(cytokine_features) <- names(cytokine_mapping)
    cytokine_features[lengths(cytokine_features) > 0]
  })
}
```

# Build the CytoSig-only databases

```{r create-db}
dbs_cytosig <- create_cytokine_dbs(dbs_all, cytokine_to_gene_mapping)

# how many cytokines survived in each database?
lengths(dbs_cytosig)
```

The resulting `dbs_cytosig` is the object used throughout the other vignettes and
in the method benchmark.

```{r session-info}
sessionInfo()
```
