---
title: "How to use CytokineFindeR"
author: "Jeffrey Tang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cytokineFindeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following is a brief introduction on the usage of `CytokineFindeR`. 

This vignette will outline the following procedures:
1) Setting up the design matrix for the linear models
2) Loading in the demo data `golimumab` from `GSE92415` on GEO. The data set is subset specifically for ulcerative colitis cell isolates looking at the gene expression difference for golimumab, which is an anti-TNF drug, between week 0 and week 6. 
3) Running the benchmark analysis in parallel using `future`
4) Data visualization which shows p-values 

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load the R packages
```{r setup}
library(cytokineFindeR)
library(tidyverse)
```

# Pull up demo data (GSE92415) which holds a cleaned up expression set and meta data subset for the treatment (golimumab) on ulcerative colitus patients
```{r}
data(golimumab)

obs_id <- golimumab$metadata$`subject:ch1`
treatment <- golimumab$metadata$`visit:ch1`

eset <- golimumab$eset
```

# Preprocess databases' receptor gene lists against the genes in the expression set of the study
# Removes ligands when no receptor genes matching the eset
```{r}
# load the default databases list of lists
dbs <- cytokineFindeR::dbs_all

# filter eset as well
preprocessed_list <- preprocess_eset(eset, dbs)

# Name variables for the preprocessed eset in the preprocessed_list
eset_f <- preprocessed_list$eset_f
dbs_f <- preprocessed_list$dbs_f
```

# Number of ligands removed:
```{r}
# baderlab: 17, nichenet: 14, fantom5: 4, citedb: 4, all_dbs: 20
dim(eset_f); sapply(dbs,length)-sapply(dbs_f, length)
```

# Set up the design matrix for demo data 
```{r}
# Input treatment: difference between week 0 and week 6
# input obs_id if paired data (in this case we do have paired data), GSM IDs different for same sample IDs due to different time points

design <- create_design(treatment, obs_id, eset_f)
design_mat <- design$design
dupcor <- design$dupcor
```

# Call the database, 
# list out the methods to benchmark, 
# run `cytokinefinder()`
# The result should return a nested list of benchmark results
```{r, warning=FALSE}
methods <- c("cfgsea", 
             "gsva_limma", 
             "pca_limma", 
             "gsva_plsda", 
             "cpca_plsda")

results <- cytokinefinder(eset_f, 
                          design_mat, 
                          dbs_f, 
                          methods, 
                          treatment = treatment,
                          obs_id = obs_id,
                          correlation = dupcor$consensus
                          )
```


# Visualize the data:
## First, extract the relevant ligands from the results object
```{r}
data_ligand <- extract_ligands(
  benchmark_results = results, 
  ligands = "TNF"
  )
plot_ligand_summary(data_ligand)
```


