---
title: "Getting started with cytokineFindeR"
author: "Jeffrey Tang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cytokineFindeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r default-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What `cytokineFindeR` does

Ligands such as cytokines are often hard to quantify directly from bulk
transcriptomics: the ligand transcript may be lowly expressed, produced by a
different cell type, or regulated post-transcriptionally. `cytokineFindeR` takes
an indirect route. For each ligand it collects the genes encoding that ligand's
**receptors** from a ligand-receptor interaction (LRI) database, treats those
receptor genes as a **gene set**, and scores how that set responds to an
experimental contrast. Ligands whose receptor sets move coherently with the
contrast are nominated as differentially active.

In short: **receptor gene set as a proxy for ligand activity.**

## The three scoring methods

A receptor gene set can be turned into a per-ligand statistic in three ways, all
exposed as functions:

- `cfgsea()` -- fgsea enrichment of the receptor set against a `limma`-ranked
  gene list (native statistic: NES).
- `gsva_limma()` -- a per-sample GSVA enrichment score for the receptor set,
  tested across the contrast with `limma` (native statistic: moderated `t`).
- `pca_limma()` -- the first principal component of the receptor set per sample,
  tested across the contrast with `limma` (native statistic: moderated `t`).

`run_lri_methods()` runs any combination of these across one or more databases
and returns a nested list of results. The companion vignette *"Comparing GSVA,
cfGSEA, and PCA for ligand inference"* benchmarks how concordant the three are
(short answer: cfGSEA and GSVA agree closely; PCA measures something different).

## The typical workflow

```{r setup}
library(cytokineFindeR)
```

The package ships a self-contained demo: `golimumab` (a cleaned anti-TNF
expression set from GSE92415, ulcerative colitis, Week 0 vs Week 6) and several
curated LRI databases.

```{r data-shapes}
data(golimumab)
data(dbs_cytosig)

## Expression matrix: genes x samples
dim(golimumab$qc_eset)

## Paired design: the same donors at two time points
table(golimumab$cond)
length(unique(golimumab$obs_id))

## Curated LRI databases (each is a named list: ligand -> receptor genes)
names(dbs_cytosig)
```

A complete analysis is four steps:

1. **Build the design.** `create_design()` constructs the contrast and, when
   `obs_id` identifies repeated donors, fits `limma::duplicateCorrelation()` for
   a paired analysis.

   ```{r design, eval = FALSE}
   design <- create_design(golimumab$cond, golimumab$obs_id, golimumab$qc_eset)
   ```

2. **Preprocess.** `preprocess_eset()` restricts the expression set to genes that
   appear in the chosen databases and drops degenerate (zero-variance) receptor
   sets.

   ```{r preprocess, eval = FALSE}
   pre <- preprocess_eset(golimumab$qc_eset, dbs_cytosig)
   ```

3. **Run the methods.** `run_lri_methods()` applies the selected methods over the
   databases.

   ```{r run, eval = FALSE}
   bench <- run_lri_methods(
     pre$eset_f, design$design, pre$dbs_f,
     methods = c("cfgsea", "gsva_limma", "pca_limma"),
     treatment = golimumab$cond, obs_id = golimumab$obs_id,
     correlation = design$dupcor$consensus.correlation
   )
   ```

4. **Summarise / rank.** `benchlist_to_tbl()` flattens the nested results into a
   tidy table and can rank a ligand of interest (e.g. TNF) across every
   method-database combination.

## Where to go next

- **End-to-end usage** with code you can run: the *"cytokineFindeR-introduction"*
  vignette.
- **Method benchmarking** (do the methods agree?): the *"Comparing GSVA, cfGSEA,
  and PCA for ligand inference"* vignette.
- **Building your own LRI gene sets** from the raw databases: the *"LRI Database
  Curation for Benchmarking"* vignette.

```{r session-info}
sessionInfo()
```
