---
title: "cytokineFindeR-introduction"
author: "Jeffrey Tang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cytokineFindeR-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The following is a brief introduction to the usage of `cytokineFindeR`.

This vignette outlines the following procedure:

1. Setting up the design matrix for the linear models.
2. Loading the demo data `golimumab` (subset from `GSE92415` on GEO). The data
   set is restricted to ulcerative colitis cell isolates and captures the gene
   expression difference for golimumab -- an anti-TNF drug -- between Week 0 and
   Week 6.
3. Running the benchmark analysis across the available methods and databases.
4. Reconstructing the nested results into a tidy table and running the CytoSig
   custom ridge model.

```{r default-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load the R package
```{r setup}
library(cytokineFindeR)
```

Load the demo data (GSE92415), which holds a cleaned-up expression set and the
metadata subset for the treatment (golimumab) in ulcerative colitis patients.
```{r load-data}
data(golimumab)

obs_id <- golimumab$obs_id
treatment <- golimumab$cond

eset <- golimumab$qc_eset
```

## Pre-filtered databases against CytoSig cytokines

The full archive of LRI reference databases is available with `data("dbs_all")`.

For this demo we use `dbs_cytosig`, a collection of LRI databases whose
ligand-receptor sets have been restricted to CytoSig cytokines (see the
*"LRI Database Curation for Benchmarking"* vignette for how it is built). It
contains several databases; here we take three source databases as an example.

```{r load-db}
# load the databases (a list of lists: database -> ligand -> receptor genes)
data("dbs_cytosig")

# pick three source databases for the demo
dbs <- dbs_cytosig[names(dbs_cytosig) %in% c("baderlab", "lianaplus", "cellchat")]
```

## Set up the design matrix for the demo data
```{r create-design_matrix}
# treatment: difference between Week 0 and Week 6.
# obs_id is supplied because the data are paired (the same donors are measured
# at both time points, with different GSM IDs), enabling a within-donor design
# via limma::duplicateCorrelation().

design <- create_design(treatment, obs_id, eset)
design_mat <- design$design
dupcor <- design$dupcor
```

## Run the preprocessing step
```{r run-preprocess}
preprocess_list <- preprocess_eset(eset, dbs)
eset_f <- preprocess_list$eset_f
dbs_f <- preprocess_list$dbs_f
```

## List the methods to benchmark and run `run_lri_methods()`

The result is a nested list of benchmark results (method -> database -> table).

```{r run-lri_methods, warning=FALSE, eval=FALSE}
methods <- c("cfgsea",
             "gsva_limma",
             "pca_limma")

results <- run_lri_methods(eset_f,
                           design_mat,
                           dbs_f,
                           methods,
                           treatment = treatment,
                           obs_id = obs_id,
                           correlation = dupcor$consensus.correlation
                          )
```

## Reconstruct the list of benchmarks into a tibble

`benchlist_to_tbl()` ranks the cytokine of interest based on the name supplied in
`golimumab_treatment`. Here the contrast targets TNF, so the function reports the
ranking of TNF across all method + database combinations.

```{r find-TNF_identification_performance, eval=FALSE}
golimumab_treatment <- list(TNF = list(benchmarks = results))
golimumab_tibble <- benchlist_to_tbl(results_list = golimumab_treatment,
                                     study_type = "treatment")
golimumab_tibble
```


## Running the CytoSig custom ridge model

The `cytosig_custom_ridge()` function runs differential expression with `limma`
and parses the logFC of all genes from the top table. The ridge regression
implementation takes the common genes based on the beta coefficients from CytoSig
and computes the cytokine signatures accordingly.

```{r run-cytosig}
data("cytosig_beta")
cytosig <- cytosig_custom_ridge(eset,
                     design_mat,
                     obs_id = obs_id,
                     correlation = dupcor$consensus.correlation,
                     beta_coef = cytosig_beta)
cytosig
```
