---
title: "Comparing GSVA, cfGSEA, and PCA for ligand inference"
author: "Jeffrey Tang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing GSVA, cfGSEA, and PCA for ligand inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r default-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

## Motivation

`cytokineFindeR` infers **ligand activity** indirectly: instead of measuring a
ligand transcript (often low or absent in bulk data), it scores the
**receptor gene set** for each ligand, taken from a ligand-receptor interaction
(LRI) database. A ligand whose receptors move coherently with the experimental
contrast is nominated as differentially active.

Three enrichment strategies are available for turning a receptor gene set into a
per-ligand statistic, and they are *not* interchangeable:

| Method       | Function       | What it scores                                  | Native statistic |
|--------------|----------------|-------------------------------------------------|------------------|
| **cfGSEA**   | `cfgsea()`     | Rank-based enrichment of the receptor set (fgsea on a limma `t` ranking) | NES |
| **GSVA**     | `gsva_limma()` | Per-sample GSVA enrichment score, then limma on the contrast | limma `t` |
| **PCA**      | `pca_limma()`  | PC1 of the receptor set per sample, then limma on the contrast | limma `t` |

This vignette benchmarks the three on a real anti-TNF dataset and asks a single,
concrete question: **do they rank ligands the same way?**

### Why rank concordance is the fair metric

The methods do not share a unit. cfGSEA reports a normalised enrichment score
(NES); GSVA and PCA report a moderated `t`-statistic from `limma`. Comparing
those numbers directly is meaningless. What *is* comparable is the **ordering**
of ligands each method produces. We therefore rank ligands within each method
(rank 1 = most up-regulated) and measure agreement with **Spearman's rho** on the
ranks. This is scale-free and answers the practical question a user actually has:
"if I trust the top of one method's list, would another method put the same
ligands there?"

```{r setup}
library(cytokineFindeR)
```

## The benchmark (pre-computed)

Running all three methods across several databases on the bundled `golimumab`
data takes a few minutes and requires the full GSVA/fgsea toolchain. To keep the
vignette fast and dependency-light, the benchmark is **run once** and its
summary tables are cached in `methods_comparison_cache.rds` (shipped with the
package). The figures below read that cache.

The chunk that follows is **shown for documentation but not executed here** -- it
is exactly the code that produced the cache. To reproduce it, set
`eval = TRUE` and supply the GSVA toolchain.

```{r heavy-compute, eval = FALSE}
## --- Reproduce the cache (run-once, ~minutes) ------------------------------
data(golimumab)
data(dbs_cytosig)

## Representative subset of the 12 CytoSig LRI databases:
## the consensus DB plus three source DBs. The full set is available in
## `dbs_cytosig` (see names(dbs_cytosig)).
reps <- c("aggregated_db", "baderlab", "lianaplus", "cellchat")
dbs  <- dbs_cytosig[reps]

eset   <- golimumab$qc_eset
cond   <- golimumab$cond        # Week 0 vs Week 6
obs_id <- golimumab$obs_id      # donor IDs -> paired design

## Filter the expression set / gene sets, build the paired design
pre  <- preprocess_eset(eset, dbs)
dsg  <- create_design(cond, obs_id = obs_id, eset = eset)
corr <- dsg$dupcor$consensus.correlation

## Run all three methods over all databases
methods <- c("cfgsea", "gsva_limma", "pca_limma")
bench <- run_lri_methods(
  eset = pre$eset_f, design = dsg$design, dbs = pre$dbs_f,
  methods = methods, treatment = cond, obs_id = obs_id,
  correlation = corr, verbose = FALSE
)

## For each method, extract the signed per-ligand effect used for ranking:
##   cfGSEA -> NES, GSVA -> t, PCA -> t  (rank 1 = largest, i.e. most up)
## then compute per-database Spearman rho between the three methods' ranks.
## (Full assembly code is in inst/scripts; the result is the cached object.)
```

```{r load-cache, echo = FALSE}
# At vignette build time the working directory is vignettes/, so the cache
# (shipped alongside this .Rmd) is found by name. After install the same file
# is copied into inst/doc/, so the relative name still resolves when the
# rendered vignette is regenerated.
cache_path <- "methods_comparison_cache.rds"
have_cache <- file.exists(cache_path)
if (have_cache) cache <- readRDS(cache_path)
have_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r cache-meta, echo = FALSE, results = "asis"}
if (have_cache) {
  m <- cache$meta
  cat(sprintf("- **Dataset:** %s\n", m$dataset))
  cat(sprintf("- **Databases:** %s\n", paste(m$databases, collapse = ", ")))
  cat(sprintf("- **Design:** %s (consensus correlation = %.2f)\n",
              m$design, m$consensus_cor))
  cat(sprintf("- **Ranking rule:** %s\n", m$rank_def))
  cat(sprintf("- **GSVA version used:** %s; built %s\n",
              m$gsva_version, m$built_on))
} else {
  cat("> Cache not found; figures are skipped. Rebuild with the chunk above.\n")
}
```

## Result 1: per-database concordance

For every database, we have three rankings of the same ligands. The table below
reports Spearman's rho for each pair of methods, per database.

```{r summary-table, echo = FALSE}
if (have_cache) {
  st <- cache$summary
  st_disp <- data.frame(
    Database          = st$database,
    `Common ligands`  = st$n_common_ligands,
    `cfGSEA~GSVA`      = round(st$rho_cfgsea_gsva, 3),
    `cfGSEA~PCA`       = round(st$rho_cfgsea_pca, 3),
    `GSVA~PCA`         = round(st$rho_gsva_pca, 3),
    check.names = FALSE
  )
  knitr::kable(st_disp, caption = "Spearman rank concordance per database.")
}
```

The pattern is consistent across all databases: **cfGSEA and GSVA agree very
strongly (rho around 0.9-0.96), while PCA agrees with neither (rho around 0).**

## Result 2: mean concordance heatmap

Averaging the pairwise rho over databases gives a compact picture of how the
three methods relate.

```{r heatmap, echo = FALSE, fig.alt = "Heatmap of mean Spearman rank concordance between the three methods."}
if (have_cache && have_ggplot) {
  ms <- cache$mean_spearman
  df <- expand.grid(method_x = rownames(ms), method_y = colnames(ms),
                    stringsAsFactors = FALSE)
  df$rho <- mapply(function(i, j) ms[i, j], df$method_x, df$method_y)
  lab <- c(cfgsea = "cfGSEA", gsva_limma = "GSVA", pca_limma = "PCA")
  df$method_x <- factor(lab[df$method_x], levels = unname(lab))
  df$method_y <- factor(lab[df$method_y], levels = rev(unname(lab)))

  ggplot2::ggplot(df, ggplot2::aes(method_x, method_y, fill = rho)) +
    ggplot2::geom_tile(color = "white", linewidth = 0.8) +
    ggplot2::geom_text(ggplot2::aes(label = sprintf("%.2f", rho)), size = 5) +
    ggplot2::scale_fill_gradient2(
      low = "#0279EE", mid = "#FAF9F3", high = "#FF9400",
      midpoint = 0, limits = c(-1, 1), name = expression(rho)) +
    ggplot2::labs(
      x = NULL, y = NULL,
      title = "Mean Spearman rank concordance across databases") +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(panel.grid = ggplot2::element_blank(),
                   axis.text = ggplot2::element_text(face = "bold"))
} else if (have_cache) {
  print(round(cache$mean_spearman, 2))
}
```

## Result 3: where the rankings agree and disagree

The scatter plots make the table concrete: each point is a ligand, positioned by
its rank under two methods. Points on the diagonal are ranked identically; a
tight diagonal cloud means high concordance, a shapeless cloud means none.

```{r scatter, echo = FALSE, fig.height = 4.2, fig.alt = "Per-ligand rank agreement between method pairs, faceted by method pair."}
if (have_cache && have_ggplot) {
  pl <- cache$pair_long
  lab <- c(cfgsea = "cfGSEA", gsva_limma = "GSVA", pca_limma = "PCA")
  pl$pair <- paste(lab[pl$method_x], "vs", lab[pl$method_y])

  ggplot2::ggplot(pl, ggplot2::aes(rank_x, rank_y)) +
    ggplot2::geom_abline(slope = 1, intercept = 0,
                         color = "grey70", linetype = 2) +
    ggplot2::geom_point(ggplot2::aes(color = database), alpha = 0.6, size = 1.4) +
    ggplot2::facet_wrap(~ pair) +
    ggplot2::scale_color_manual(
      values = c(aggregated_db = "#000000", baderlab = "#FF9400",
                 lianaplus = "#75A025", cellchat = "#0279EE"),
      name = "Database") +
    ggplot2::labs(
      x = "Rank (method on x)", y = "Rank (method on y)",
      title = "Per-ligand rank agreement between methods") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(legend.position = "bottom")
} else if (have_cache) {
  cat("Install ggplot2 to render the rank-agreement scatter plots.\n")
}
```

The **cfGSEA vs GSVA** panel hugs the diagonal; the two PCA panels are diffuse.

## Interpretation

This divergence is expected once you look at *what each method measures*:

- **cfGSEA and GSVA both summarise the receptor set as a whole.** cfGSEA asks
  whether the receptor genes are enriched toward the top of a differential
  ranking; GSVA builds a per-sample enrichment score from the same set and tests
  it. Both are, in effect, **"is this set coherently up or down?"** statistics, so
  they order ligands almost identically.

- **PCA scores a different quantity.** `pca_limma()` reduces the receptor set to
  its first principal component -- the dominant **axis of covariation** among the
  receptors -- and tests whether *that axis* shifts with the contrast. PC1 can be
  driven by a few high-variance receptors and need not point in the direction of
  mean enrichment. It can even be sign-arbitrary. So PCA answers a related but
  distinct question and produces a near-orthogonal ranking.

**Practical guidance.** If the goal is to nominate ligands whose receptor sets
move coherently with a treatment (the usual `cytokineFindeR` use case), cfGSEA
and GSVA are mutually corroborating and either is a reasonable default. PCA is
best read as a complementary view that flags ligands whose receptors have strong
*structured covariation* -- useful, but it should not be expected to agree with
the enrichment-based methods, and its results warrant separate interpretation.

## Session information

```{r session-info}
sessionInfo()
```
